A highly-available move operation for replicated trees

### 2019-06-05 paper/move-op.tex

CRDTs rock \cite{Shapiro:2011wy,Shapiro:2011un}.

>>>>

For many systems and applications, the data model that describes the system state is a tree.
For example:
\begin{itemize}
    \item Most filesystems, e.g. on Windows and Unix-like OSes, are organised as a tree: folders or directories are branch nodes, and files are leaf nodes.\footnote{Although symbolic links appear to allow non-tree graph structures, such as cyclic references, the underlying inode structure remains a tree: a symbolic link is not a reference to another inode, but rather a special file containing the path of another file or directory, which is resolved to an inode when the symlink is accessed.
        If POSIX hardlinks are allowed, the filesystem is a restricted DAG rather than a tree, because the same file inode can be referenced from multiple places in the directory tree.
        However, hardlinks are not supported by Windows, nor by popular file sync products such as Dropbox, Google Drive or Microsoft OneDrive, so we do not consider hardlinks further in this paper.}
        In Unix-style filesystems, the nodes of the tree are known as \emph{inodes}.
    \item Rich text editors maintain a tree of textual structures such as paragraphs, bulleted and numbered lists, figures, sections, and so on.
        These can be nested inside each other: for example, a paragraph may appear inside a bullet point of a bulleted list, which may in turn appear inside a numbered point of an enumeration.
        Rich text documents are typically represented in memory using the Document Object Model (DOM), and written to file in HTML, XML, or JSON format.
    \item Vector graphics and presentation software represents images using graphical objects such as text boxes, rectangles, ellipses, lines, and so on.
        These objects are contained within nodes representing pages of a document, or slides in a presentation.
        Multiple objects may be combined in a \emph{group} so that they can be moved or manipulated as a unit; this corresponds to making these objects children of a common parent node.
        Multiple objects and groups may in turn be combined into higher-level groups, forming a tree.
        Moreover, the order of nodes determines visibility (i.e.\ objects that are ``further back'' may be obscured by objects that are ``closer to the front'').
    \item Personal note-taking and task-management tools such as org-mode for Emacs, or Workflowy, explicitly present the user with a tree structure that they can inspect and manipulate.
\end{itemize}

In this paper we consider applications that use such a tree data model, and that replicate this tree across multiple nodes.
Moreover, we focus on \emph{optimistic} replication~\cite{Saito:2005jw}, that is, systems in which any replica can autonomously make changes to the data, without waiting for communication or coordination with any other replicas.
Such systems have the advantage that they can continue processing read and write requests even in the presence of arbitrary network partitions; in other words, they are \emph{available} and \emph{partition-tolerant} in the sense of the CAP theorem~\cite{Gilbert:2002il}.

CRDTs~\cite{Shapiro:2011wy,Shapiro:2011un}

\section{Why a move operation is hard}

Errors in Google Drive

% - No Synchronisation or consensus required. Available in the sense of CAP
% - Rely only on uniqueness of node IDs, e.g. inode numbers
% - Composing JSON/XML tree by using list datatype
% - Location is Lww register. Concurrent move does not duplicate
% - Isabelle: combines the conciseness of pseudocode with the precision of mathematical notation.
% - Unlike NFS, can work offline
